---
title: "COVID-19 Map"
author: |
  | Yanchang Zhao
  | yanchang@RDataMining.com
  | http://RDataMining.com
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, echo=T, cache=F, warning=F, message=F, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr) # pipe operations
library(lubridate) # date operations
library(tidyverse) # ggplot2, tidyr, dplyr...
library(leaflet) # map
```



This report presents a world map of the Novel Coronavirus (COVID-19), using R and *leaflet*.

The data source used for this analysis is *the 2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository*^[https://github.com/CSSEGISandData/COVID-19] built by the Center for Systems Science and Engineering, Johns Hopkins University.

The map shows the number of confirmed cases around the world. Click on the circles on the map to see the number of confirmed cases.

```{r pressure, echo=FALSE}
library(magrittr) # pipe operations
library(lubridate) # date operations
library(tidyverse) # ggplot2, tidyr, dplyr...
library(leaflet) # map

## source data files
# filenames <- c('time_series_covid19_confirmed_global.csv', 
#                'time_series_covid19_deaths_global.csv')
url.path <- paste0('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/',
                   'master/csse_covid_19_data/csse_covid_19_time_series/')

## download files to local
# download <- function(filename) {
#   url <- file.path(url.path, filename)
#   dest <- file.path('./data', filename)
#   download.file(url, dest)
# }
# bin <- lapply(filenames, download)

## load data into R
data.confirmed <- read.csv(file.path(url.path, 'time_series_covid19_confirmed_global.csv'))
# data.confirmed <- read.csv('./data/time_series_covid19_confirmed_global.csv') 
# data.deaths <- read.csv('./data/time_series_covid19_deaths_global.csv') 

# latest date
x <- data.confirmed
max.date <- names(x)[ncol(x)] %>% substr(2,8) %>% mdy()
max.date.txt <- max.date %>% format('%d %B %Y')


## select last column, which is the number of latest confirmed cases
x$confirmed <- x[, ncol(x)]
x %<>% select(c(Country.Region, Province.State, Lat, Long, confirmed)) %>%
  mutate(txt=paste0(Country.Region, ' - ', Province.State, ': ', confirmed))

m <- leaflet(width=1200, height=800) %>% addTiles()
# circle marker
m %>% addCircleMarkers(x$Long, x$Lat, 
                       #radius=log2(x$confirmed), stroke=F,
                       radius=0.03*sqrt(x$confirmed), stroke=F,
                       color='red', fillOpacity=0.5,
                       popup=x$txt)

```

Created with the latest data on `r max.date.txt` UTC.
